---
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

---
title: "OLS Regression of Charitable Contributions on Income"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r include=FALSE}
# Install and load necessary libraries
if(!require(dplyr)){install.packages("dplyr")}
if(!require(lmtest)){install.packages("lmtest")}
if(!require(sandwich)){install.packages("sandwich")}
if(!require(stargazer)){install.packages("stargazer")}
if(!require(Metrics)){install.packages("Metrics")}

library(dplyr)
library(lmtest)
library(sandwich)
library(stargazer)
library(Metrics)
library(ggplot2); theme_set(theme_classic())
library(knitr)
library(tidyverse)
library(scales)
library(broom)
```

```{r include=FALSE}
df <- read.csv("../data/interim/interim_county_irs_grouped.csv")
```

### Data Splitting

```{r include=FALSE}
# Set a seed for reproducibility
set.seed(123)

# Determine the number of rows
n_rows <- nrow(df)
# Calculate the size of the exploration and confirmation sets (30%/70%)
exploration_size <- floor(0.3 * n_rows)
confirmation_size <- n_rows - exploration_size


# Randomly sample indices for the exploration set
exploration_indices <- sample(1:n_rows, size = exploration_size, replace = FALSE)

# Create the exploration and confirmation data frames
exploration_df <- df[exploration_indices, ]
confirmation_df <- df[-exploration_indices, ]

nrow(confirmation_df)


```

### EDA

```{r histograms}
exploration_df %>%
  ggplot(aes(x = mean_charitable_contributation_per_return)) +
  geom_histogram(aes(y = stat(count/sum(count)))) +
  labs(x = "Mean Charitable Contribution per Return", y = "Proportion",
       title = "Histogram of Mean Charitable Contribution per Return")

exploration_df %>%
  ggplot(aes(x = mean_agi_per_return)) +
  geom_histogram(aes(y = stat(count/sum(count)))) +
  labs(x = "Mean AGI per Return", y = "Proportion",
       title = "Histogram of Mean AGI per Return")


exploration_df %>%
  ggplot(aes(x = mean_agi_per_return, y = mean_charitable_contributation_per_return)) +
  geom_point() + 
  geom_smooth(method = "lm")

exploration_df %>%
  ggplot(aes(x = mean_agi_per_return, y = mean_charitable_contributation_per_return)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))

exploration_df %>%
  ggplot(aes(x = log(mean_agi_per_return))) +
  geom_histogram()

exploration_df %>%
  ggplot(aes(x = mean_dividend_per_dividend_return)) +
  geom_histogram()


exploration_df %>%
  ggplot(aes(x = mean_capgain_per_capgain_return)) +
  geom_histogram()

exploration_df %>%
  ggplot(aes(x = mean_rent_royalty_per_rent_royalty_return)) +
  geom_histogram()


exploration_df %>%
  ggplot(aes(x = mean_salary_per_salary_return)) +
  geom_histogram()



exploration_df %>%
  ggplot(aes(x = total_net_investment_income_tax_amount)) +
  geom_histogram()


exploration_df %>%
  ggplot(aes(x = mean_tax_payment_per_return)) +
  geom_histogram()

```

### Operationalization

The selection and transformation of variables are crucial steps in building a robust and interpretable model. We operationalize the key concepts from the literature on charitable giving as follows. Income is represented by mean_agi_per_return, which captures the average adjusted gross income per tax return. To account for potential non-linear effects of income, we include both a quadratic term (mean_agi_per_return squared) and logarithmic transformations (natural logarithm of mean_agi_per_return and its squared term). Wealth and the availability of discretionary funds are operationalized using mean_dividend_per_dividend_return, mean_capgain_per_capgain_return, and mean_rent_royalty_per_rent_royalty_return, representing average income from dividends, capital gains, and rent/royalties, respectively. Stable income sources are captured by mean_salary_per_salary_return, the average salary per tax return. Finally, to operationalize the influence of tax incentives, we include total_net_investment_income_tax_amount and mean_tax_payment_per_return.

This approach translates the theoretical concepts of income, wealth, and tax effects into specific, measurable variables. The transformations applied to mean_agi_per_return are designed to address its potentially skewed distribution and to capture non-linear relationships with charitable giving. For instance, the quadratic term allows us to model a potential U-shaped or inverted U-shaped relationship, while the logarithmic transformation can help normalize the data and allows us to interpret the coefficient on income as an elasticity.

<!-- #region -->
### Model Specifications
We constructed a sequence of 8 regression models including 4 main ones, incrementally adding theoretically relevant predictor variables to assess their impact on the outcome variable with different transformations. This sequential approach allows us to systematically examine the influence of different economic factors on philanthropic behavior per tax return. Model 1 examines a simple linear relationship. Model 2 introduces a quadratic term for mean_agi to capture potential non-linear effects. Model 3 explores a similar non-linear relationship using the natural logarithm of mean_agi and its squared term. Models 4 incorporate variables that proxy for wealth and the availability of discretionary funds. Model 5 addtionally have stable income sources. Model 6 &7 add tax-related variables. Models 8 is again using the log transformation with adding variables that proxy for wealth and the availability of discretionary funds.


### Model Building
<!-- #endregion -->

```{r include=FALSE}
# Model 1: Simple OLS (Income)
model1 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return, data = confirmation_df)
#robust_summary_model1 <- coeftest(model1, vcov = vcovHC(model1, type = "HC3"))
robust_summary_model1 <- vcovHC(model1, type = "HC3")
#print("Summary of Model 1 (Simple OLS with Robust SE):")
#print(summary(model1))
#print("\nCoefficients (with Robust Standard Errors) - Model 1:")
#print(robust_summary_model1)
#cat("\n")
```

```{r include=FALSE}
# Model 2: Polynomial Income (Quadratic)
model2 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2), data = confirmation_df)
robust_summary_model2 <- vcovHC(model1, type = "HC3")
#print("Summary of Model 2 (Polynomial Income with Robust SE):")
#print(summary(model2))
#print("\nCoefficients (with Robust Standard Errors) - Model 2:")
#print(robust_summary_model2)
#cat("\n")
```

```{r include=FALSE}
# Model 3: Log Income with Quadratic Term
model3 <- lm(mean_charitable_contributation_per_return ~ log(mean_agi_per_return) + I(log(mean_agi_per_return)^2), data = confirmation_df)
robust_summary_model3 <- coeftest(model3, vcov = vcovHC(model3, type = "HC3"))
# print("Summary of Model 3 (Log Income with Quadratic Term with Robust SE):")
# print(summary(model3))
# print("\nCoefficients (with Robust Standard Errors) - Model 3:")
# print(robust_summary_model3)
# cat("\n")
```

```{r include=FALSE}
# Model 4: Adding Wealth/Investment Income
model4 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return, data = confirmation_df)
robust_summary_model4 <- coeftest(model4, vcov = vcovHC(model4, type = "HC3"))
# print("Summary of Model 4 (Adding Wealth/Investment Income with Robust SE):")
# print(summary(model4))
# print("\nCoefficients (with Robust Standard Errors) - Model 4:")
# print(robust_summary_model4)
# cat("\n")
```

```{r include=FALSE}
# Model 5: Adding Stable Income (Salary)
model5 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return, data = confirmation_df)
robust_summary_model5 <- coeftest(model5, vcov = vcovHC(model5, type = "HC3"))
# print("Summary of Model 5 (Adding Stable Income with Robust SE):")
# print(summary(model5))
# print("\nCoefficients (with Robust Standard Errors) - Model 5:")
# print(robust_summary_model5)
# cat("\n")
```

```{r include=FALSE}
# Model 6: Adding Tax-Related Variable (Investment Income Tax)
model6 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return + total_net_investment_income_tax_amount, data = confirmation_df)
robust_summary_model6 <- vcovHC(model1, type = "HC3")
#model6
#print("Summary of Model 6 (Adding Investment Income Tax with Robust SE):")
#print(summary(model6))
#print("\nCoefficients (with Robust Standard Errors) - Model 6:")
#print(robust_summary_model6)
#cat("\n")
```

```{r include=FALSE}
# Model 7: Adding Another Tax-Related Variable (Tax Payment)
model7 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return + total_net_investment_income_tax_amount + mean_tax_payment_per_return, data = confirmation_df)
robust_summary_model7 <- coeftest(model7, vcov = vcovHC(model7, type = "HC3"))
# print("Summary of Model 7 (Adding Tax Payment with Robust SE):")
# print(summary(model7))
# print("\nCoefficients (with Robust Standard Errors) - Model 7:")
# print(robust_summary_model7)
# cat("\n")
```

```{r include=FALSE}
# Model 8: Log Income with Quadratic Term and Wealth/Investment Income
model8 <- lm(mean_charitable_contributation_per_return ~ log(mean_agi_per_return) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return, data = confirmation_df)
robust_summary_model8 <- coeftest(model8, vcov = vcovHC(model8, type = "HC3"))
# print("Summary of Model 8 (Log Income and Wealth/Investment Income with Robust SE):")
# print(summary(model8))
# print("\nCoefficients (with Robust Standard Errors) - Model 8:")
# print(robust_summary_model8)
# cat("\n")
```

```{r}
# Model 9: Log Income with Quadratic Term and All Other Variables
model9 <- lm(mean_charitable_contributation_per_return ~ log(mean_agi_per_return) + I(log(mean_agi_per_return)^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return + total_net_investment_income_tax_amount + mean_tax_payment_per_return, data = confirmation_df)
robust_summary_model9 <- coeftest(model9, vcov = vcovHC(model9, type = "HC3"))
robust_summary_model9
# print("Summary of Model 9 (Log Income and All Other Variables with Robust SE):")
# print(summary(model9))
# print("\nCoefficients (with Robust Standard Errors) - Model 9:")
# print(robust_summary_model9)
# cat("\n")
```

### Assumptions (plots can be in the appendix for models 4 & 8 maybe?)

<!-- #region -->
#### IID



#### BLP Exist

The BLP, by definition, assumes a linear relationship between the
dependent variable and the independent variables. We can assess the
appropriateness of this assumption by:

Scatter Plots: Visualizing the relationship between each independent
variable and the dependent variable using scatter plots. Look for
non-linear patterns.


Residual Analysis: After estimating a linear model, examine the
residuals for non-linear patterns. For example, a scatter plot of
residuals against predicted values might reveal a curved pattern.

<!-- #endregion -->

```{r}
# Create a list of your models
model_list <- list(
  "Model 1" = model1,
  "Model 2" = model2,
  "Model 3" = model3,
  "Model 4" = model4,
  "Model 5" = model5,
  "Model 6" = model6,
  "Model 7" = model7,
  "Model 8" = model8,
  "Model 9" = model9
)
```

#### Assessing Perfect Collinearity 

```{r}
library(car)
```

```{r}


# Start loop from i = 1
for (i in 2:length(model_list)) {
  model <- model_list[[i]]
  
  cat("\n--- VIF for Model", i, "---\n")
  print(vif(model))
}
```

#### Linear Conditional Expectation 

```{r}
for (i in seq_along(model_list)) {
  model <- model_list[[i]]
  stud_res <- rstudent(model)
  fitted_vals <- fitted(model)
  
  # Plot: Studentized Residuals vs Fitted Values
  plot(fitted_vals, stud_res,
       main = paste("Studentized Residuals vs Fitted - Model", i),
       xlab = "Fitted Values",
       ylab = "Studentized Residuals",
       pch = 19, col = "darkblue")
  
  # Reference line at 0
  abline(h = 0, col = "red", lwd = 2, lty = 2)
  
}
```

#### Normal Distribution of Errors

```{r}
for (i in seq_along(model_list)) {
  model <- model_list[[i]]
  stud_res <- rstudent(model)
  
  # Set up side-by-side plots
  par(mfrow = c(1, 2))
  
  # Q-Q plot of studentized residuals
  qqnorm(stud_res,
         main = paste("Q-Q Plot (Studentized) - Model", i),
         pch = 19, col = "darkblue")
  qqline(stud_res, col = "red", lwd = 2)
  
  # Histogram as density plot
  hist(
    stud_res,
    main = paste("Density Histogram - Model", i),
    xlab = "Studentized Residuals",
    col = "lightcoral",
    border = "white",
    freq = FALSE
  )
  
  # Overlay a normal density curve
  curve(dnorm(x, mean = mean(stud_res), sd = sd(stud_res)),
        col = "blue", lwd = 2, add = TRUE)
  
  # Reset plotting area
  par(mfrow = c(1, 1))
  
}

```

#### Homoskedasticity

```{r}


# --- Step 1: Run Breusch-Pagan Test for all models and create a table ---

bp_results_list <- list()
if (exists("model_list") && length(model_list) > 0)
{
  for (i in 1:length(model_list)) {
    model_name <- names(model_list)[i]
    if (is.null(model_name)) {
      model_name <- paste("Model", i)
    }
    current_model <- model_list[[i]]

    # Perform Breusch-Pagan test
    bp_test <- bptest(current_model)

    # Tidy the result and add model name
    bp_results_list[[model_name]] <- tidy(bp_test) %>%
      mutate(Model = model_name) %>%
      select(Model, statistic, p.value) # Select relevant columns
  }

  # Combine results into a single data frame
  bp_results_table <- bind_rows(bp_results_list) %>%
    mutate(p.value_formatted = format(p.value, scientific = TRUE, digits = 3))


  print("Final Breusch-Pagan Test Results Table:")
  # Print the table using kable
  print(
    kable(
      bp_results_table %>% select(Model, statistic, p.value_formatted),
      col.names = c("Model", "BP Statistic", "p-value"),
      caption = "Breusch-Pagan Test Results for Homoskedasticity"
    )
  )

} else {
print("No successful Breusch-Pagan test results were collected.")
}
## visualize the p-values from the BP test table
 {

  # Define the significance level
  alpha <- 0.05

  plot_bp_pvalues_log <- ggplot(bp_results_table, aes(x = reorder(Model, p.value), y = p.value)) +
    geom_col(fill = "skyblue") +
    # Add significance line - use log10(alpha) for its position on the log scale
    geom_hline(yintercept = alpha,
               linetype = "dashed",
               color = "red") +
    scale_y_log10(
      name = "p-value (log10 scale)",
      breaks = 10^seq(-150, 0, by = 10),
      labels = scales::trans_format("log10", scales::math_format(10^.x))
    ) +
    annotation_logticks(sides = "l") + # Add log tick marks on the left axis
    labs(
      title = "Breusch-Pagan Test p-values by Model (Log Scale)",
      subtitle = paste(
        "Lower p-value indicates significant heteroskedasticity (Bars crossing red line where p <",
        alpha,
        ")"
      ),
      x = "Model",
      y = "p-value" # Corrected y-axis label to match the actual data
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels if needed

  print(plot_bp_pvalues_log)

} 
```

#### Assumptions Summary


We selected Model 4 and Model 8 for their favorable multicollinearity scores, showing no significant issues. Model 4 uses a polynomial transformation of mean_agi, while Model 8 employs a log transformation of mean_agi. Both models include key predictors like mean_dividend, mean_capgain, and mean_rent_royalty, with no problematic VIFs. We excluded stable income variables and tax-related predictors, as they showed collinearity with the outcome. These two models effectively capture the relationships without multicollinearity concerns.

Plots for Model 4 showed heteroscedasticity and non-normal residuals (lighter lower, heavier upper tails). The Breusch-Pagan test for Model 4 was significant (p < 0.001), indicating non-constant variance. Model 8 plots suggested better variance consistency but still showed non-normal residuals (prominent heavier upper tail). Crucially, the Breusch-Pagan test for Model 8 was even more significant (p < 0.001, BP statistic much higher than Model 4), indicating a more severe issue of non-constant variance compared to Model 4, despite the visual assessment of the residual plot. Both models also exhibit non-normality in their residual distributions.

```{r include=FALSE}
se1 <- vcovHC(model1, type = "HC3")
rse1 <- sqrt(diag(se1))

se2 <- vcovHC(model2, type = "HC3")
rse2 <- sqrt(diag(se2))

se3 <- vcovHC(model3, type = "HC3")
rse3 <- sqrt(diag(se3))

se4 <- vcovHC(model4, type = "HC3")
rse4 <- sqrt(diag(se4))

se5 <- vcovHC(model5, type = "HC3")
rse5 <- sqrt(diag(se5))

se6 <- vcovHC(model6, type = "HC3")
rse6 <- sqrt(diag(se6))

se7 <- vcovHC(model7, type = "HC3")
rse7 <- sqrt(diag(se7))

se8 <- vcovHC(model8, type = "HC3")
rse8 <- sqrt(diag(se8))

se9 <- vcovHC(model9, type = "HC3")
rse9 <- sqrt(diag(se9))
```

```{r building stargazer}
m1 <- model1
m2 <- model2
m3 <- model3
m4 <- model4
m5 <- model5
m6 <- model6
m7 <- model7
m8 <- model8
m9 <- model9

show_F_in_two_lines <- function(stargazer) {
  # `Stringr` works better than base's regex 
  require(stringr)

  # If you remove `capture.output()`, not only the modified LaTeX code 
  # but also the original code would show up
  stargazer <- stargazer |>
    capture.output()

  # Reuse the index in which F-statistics are displayed
  position_F <- str_which(stargazer, "F Statistic")

  # Extract only F-statistics
  Fs <- stargazer[position_F] |>
    str_replace_all("\\(.*?\\)", "")

  # Extract only df values and make a new line for them
  dfs <- stargazer[position_F] |>
    str_extract_all("\\(.*?\\)") |>
    unlist() |>
    (
      \(dfs)
      paste0(" & ", dfs, collapse = "")
    )() |>
    paste0(" \\\\")

  # Reuse table elements that are specified
  # after the index of F-statistics
  after_Fs <- stargazer[-seq_len(position_F)]

  c(
    stargazer[seq_len(position_F - 1)],
    Fs,
    dfs,
    after_Fs
  ) |>
    cat(sep = "\n")
}

summary(m5)
cov_labels <- c("mean AGI", "mean AGI**2", "log(mean AGI)", 
                "log(mean AGI**2", "mean dividend", "mean capgain", "mean rent royalty",
                "mean salary",  "Intercept")


cov_labels2 <- c("mean AGI", "mean AGI**2", "log(mean AGI)", 
                "log(mean AGI**2", "mean dividend", "mean capgain", "mean rent royalty",
                "mean salary", "net investment income","total net investment income tax amount", "mean tax payment", "Intercept")

stargazer(m1, m2, m3, m4, m5, type = "latex",
          out = "stargazertable.tex",
          se = list(se1, se2, se3, se4, se5),
          header = F,
          font.size = "tiny",
          table.placement = "t",
          column.sep.width = ".001pt",
          omit.stat = c("ser", "rsq"),
          covariate.labels = cov_labels,
          dep.var.caption = "Dependent Variable: mean charitable contribution",
          dep.var.labels.include = F,
          float = F,
          digits = 3,
          notes = c("Robust standard errors are displayed.", "The covariate names omit per unit (e.g., per return) due to space limit.")) %>%
  show_F_in_two_lines()
  

stargazer(m1, m6, m7, m8, m9, type = "latex",
          se = list(se1, se6, se7, se8, se9),
          header = F,
          font.size = "tiny",
          table.placement = "b",
          column.sep.width = ".001pt",
          column.labels = c("(1)", "(6)", "(7)", "(8)", "(9)"),
          model.numbers = F,
          omit.stat = c("ser", "rsq"),
          covariate.labels = cov_labels2,
          dep.var.caption = "Dependent Variable: mean charitable contribution",
          dep.var.labels.include = F,
          float = F,
          digits = 3,
          notes = c("Robust standard errors are displayed.", "The covariate names omit per unit (e.g., per return) due to space limit.")) %>%
  show_F_in_two_lines()


# are we going to use rsq or adjusted rsq?
```

### Results & Discussion

<!-- #region -->
All models significantly predict mean charitable contributions (p < 0.001). Models incorporating multiple income sources generally exhibit higher explanatory power compared to the baseline models relying solely on AGI. Models 4 and 5, including wealth/investment income and salary, show higher explanatory power than Models 1 and 2 (which only include AGI and AGI squared). Specifically, Model 1 (mean AGI) explains 60.2% of the variance, while Model 2 (with AGI squared) increases this to 70.6%. Adding wealth/investment income (Model 4) further increases it to 78.5%. This suggests income, wealth/investment, significantly influence charitable giving. Models 5, 6, and 7 after including salary and tax-related variables reached 79.0%-79.2%. However, these exhibit significant multicollinearity as shown by their high VIF scores. Model 8 shows a 72.1% of adjusted R-squared. After assessing all of the linear regression assumptions and performance, we believe Model 4 may be a more suitable choice due to its balance of strong explanatory power and lesser degree of assumptions violations.

Here is a practical example using model 4's coefficients:
For every \\$1,000 increase in mean dividend income, charitable contributions are predicted to increase by \\$49, suggesting individuals with higher dividend earnings are, on average, expected to donate more, highlighting the influence of investment income on charitable giving. 
###### [To be filled: Impact on AGI]


In a broader sense, these findings indicate that charitable giving is a complex behavior influ
enced by a variety of predictors. Stable income, such as salary, and income from investments,
 such as dividends and capital gains, both play a role. The results corroborate our initial hy
pothesis of a non-linear, U-shaped relationship between AGI and charitable contributions. These
 results may be useful for charitable organizations and researchers exploring complex dynamics
 associated with charitable behaviors.
<!-- #endregion -->

### Results & Discussion for model 5 as the best (Previous results)

The models significantly predict mean charitable contributions (p < 2.2e-16), with adjusted R-squared values indicating varying explanatory power. Models 4 and 5, including wealth/investment income and salary, show higher explanatory power than Models 1 and 2 (which only include AGI and AGI squared). Specifically, Model 1 (mean AGI) explains 60.2%, while Model 2 (with AGI squared) increases this to 70.6%. Adding wealth/investment income (Model 4) further increases it to 78.5%, and including salary (Model 5) reaches 79.0%. This suggests income, wealth/investment, and salary significantly influence charitable giving. While Models 6 and 7 exhibit slightly higher R-squared values shown in the appendix, Model 5 may be a more suitable choice due to its balance of strong explanatory power and greater parsimony.

A $1,000 increase in mean salary is predicted to increase charitable contributions by $12, holding other variables constant. This suggests that a $10,000 increase in salary would lead to an additional $120 in predicted charitable giving, indicating a positive relationship between salary and charitable donations. For every $1,000 increase in mean dividend income, charitable contributions are predicted to increase by $54, suggesting individuals with higher dividend earnings are, on average, expected to donate more, highlighting the influence of investment income on charitable giving. The model captures a U-shaped relationship between adjusted gross income (AGI) and charitable donations. At very low AGI levels, say around $15,000, individuals may have significant financial constraints. As their income slightly rises to $25,000, they might still be focused on meeting basic needs, and charitable giving might remain relatively low or even decrease as they have less disposable income after taking on new expenses. However, as income continues to increase to $50,000 or $75,000, individuals might have more discretionary income, and their charitable giving could increase at an increasing rate. This is because the initial negative effect of AGI on giving is eventually overcome by the positive effect of the squared AGI term as income grows. 

 In a broader sense, these findings indicate that charitable giving is a complex behavior influ
enced by a combination of factors. Stable income, such as salary, and income from investments,
 such as dividends and capital gains, both play a role. The results corroborate our initial hy
pothesis of non-linear, U-shaped relationship between AGI and charitable contributions. These
 results may be useful for charitable organizations and researchers exploring complex dynamics
 associated with charitable behaviors.

```{r}

```
