---
output:
  pdf_document: default
  html_document: default
---


---
title: "OLS Regression of Charitable Contributions on Income"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r include=FALSE}
# Install and load necessary libraries
if(!require(dplyr)){install.packages("dplyr")}
if(!require(lmtest)){install.packages("lmtest")}
if(!require(sandwich)){install.packages("sandwich")}
if(!require(stargazer)){install.packages("stargazer")}
if(!require(Metrics)){install.packages("Metrics")}

library(dplyr)
library(lmtest)
library(sandwich)
library(stargazer)
library(Metrics)
```

```{r include=FALSE}
df <- read.csv("../data/interim/interim_irs_county.csv")
```

### Methodology

To investigate the factors influencing mean charitable contributions per tax return, our methodology involves employing Ordinary Least Squares (OLS) regression on a dataset that has been split into training (80%) and testing (20%) sets. The model building process will be conducted exclusively on the training data to develop predictive relationships, while the unseen testing data will be used for subsequent model evaluation. We will construct a sequence of nine regression models, incrementally adding theoretically relevant predictor variables to assess their impact on the outcome variable, `mean_charitable_contributation_per_return`. Given the established literature on charitable giving, this sequential approach allows us to systematically examine the influence of different economic factors on philanthropic behavior per tax return. To address potential issues of heteroskedasticity, robust standard errors will be calculated for the coefficients of all estimated models, providing more reliable statistical inference.

Our model building process begins by establishing the core relationship between income, as measured by `mean_agi_per_return`, and charitable contributions through three initial specifications. Model 1 examines a simple linear relationship. Model 2 introduces a quadratic term for `mean_agi_per_return` to capture potential non-linear effects, such as diminishing or increasing returns to income in charitable giving. Model 3 explores a similar non-linear relationship using the natural logarithm of `mean_agi_per_return` and its squared term, which can be particularly useful given the often skewed distribution of income. Following this, Model 4 incorporates variables that proxy for wealth and the availability of discretionary funds: `mean_dividend_per_dividend_return`, `mean_capgain_per_capgain_return`, and `mean_rent_royalty_per_rent_royalty_return`, based on the hypothesis that individuals with higher levels of investment income and income from assets have a greater capacity for charitable giving. Model 5 then includes `mean_salary_per_salary_return` to examine the independent effect of stable income sources after controlling for overall income and wealth-related income. Finally, Models 6 and 7 incorporate tax-related variables, `total_net_investment_income_tax_amount` and `mean_tax_payment_per_return`, to reflect the potential influence of tax incentives and the benefit of charitable deductions. Models 8 and 9 revisit the use of log-transformed AGI with a quadratic term, both with and without the full set of control variables, to compare the model fit and significance of predictors under this alternative functional form of income. By examining the changes in model fit and the significance of coefficients across these nine specifications on the training data, and subsequently evaluating their predictive performance on the held-out testing data, we aim to provide a comprehensive analysis of the economic factors influencing mean charitable contributions per tax return.


The below need to be turned into a table 

**Model 1 (Linear Income):**
The mean charitable contribution per return is modeled as a linear function of the mean adjusted gross income (AGI) per return, plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \text{mean\_agi\_per\_return}_i + \epsilon_i $$

**Model 2 (Polynomial Income - Quadratic):**
The mean charitable contribution per return is modeled as a quadratic function of the mean AGI per return, plus an error term. This includes both a linear and a squared term for AGI.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \text{mean\_agi\_per\_return}_i + \beta_2 \cdot (\text{mean\_agi\_per\_return}_i)^2 + \epsilon_i $$

**Model 3 (Log Income with Quadratic Term):**
The mean charitable contribution per return is modeled as a quadratic function of the natural logarithm of the mean AGI per return, plus an error term. This includes both a linear and a squared term for the log of AGI.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \ln(\text{mean\_agi\_per\_return}_i) + \beta_2 \cdot (\ln(\text{mean\_agi\_per\_return}_i))^2 + \epsilon_i $$

**Model 4 (Polynomial Income + Wealth):**
The mean charitable contribution per return is modeled as a function of the mean AGI per return (linear and quadratic terms) and several wealth/investment income variables: mean dividend per dividend return, mean capital gain per capital gain return, and mean rent/royalty per rent/royalty return, plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \text{mean\_agi\_per\_return}_i + \beta_2 \cdot (\text{mean\_agi\_per\_return}_i)^2 + \beta_3 \cdot \text{mean\_dividend\_per\_dividend\_return}_i + \beta_4 \cdot \text{mean\_capgain\_per\_capgain\_return}_i + \beta_5 \cdot \text{mean\_rent\_royalty\_per\_rent\_royalty\_return}_i + \epsilon_i $$

**Model 5 (Polynomial Income + Wealth + Salary):**
The mean charitable contribution per return is modeled as a function of the mean AGI per return (linear and quadratic terms), the wealth/investment income variables (as in Model 4), and the mean salary per salary return, plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \text{mean\_agi\_per\_return}_i + \beta_2 \cdot (\text{mean\_agi\_per\_return}_i)^2 + \beta_3 \cdot \text{mean\_dividend\_per\_dividend\_return}_i + \beta_4 \cdot \text{mean\_capgain\_per\_capgain\_return}_i + \beta_5 \cdot \text{mean\_rent\_royalty\_per\_rent\_royalty\_return}_i + \beta_6 \cdot \text{mean\_salary\_per\_salary\_return}_i + \epsilon_i $$

**Model 6 (Polynomial Income + Wealth + Salary + NIIT):**
The mean charitable contribution per return is modeled as a function of the mean AGI per return (linear and quadratic terms), the wealth/investment income variables, the mean salary per salary return, and the total net investment income tax amount, plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \text{mean\_agi\_per\_return}_i + \beta_2 \cdot (\text{mean\_agi\_per\_return}_i)^2 + \beta_3 \cdot \text{mean\_dividend\_per\_dividend\_return}_i + \beta_4 \cdot \text{mean\_capgain\_per\_capgain\_return}_i + \beta_5 \cdot \text{mean\_rent\_royalty\_per\_rent\_royalty\_return}_i + \beta_6 \cdot \text{mean\_salary\_per\_salary\_return}_i + \beta_7 \cdot \text{total\_net\_investment\_income\_tax\_amount}_i + \epsilon_i $$

**Model 7 (Polynomial Income + Wealth + Salary + NIIT + MTAX):**
The mean charitable contribution per return is modeled as a function of the mean AGI per return (linear and quadratic terms), the wealth/investment income variables, the mean salary per salary return, the total net investment income tax amount, and the mean tax payment per return, plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \text{mean\_agi\_per\_return}_i + \beta_2 \cdot (\text{mean\_agi\_per\_return}_i)^2 + \beta_3 \cdot \text{mean\_dividend\_per\_dividend\_return}_i + \beta_4 \cdot \text{mean\_capgain\_per\_capgain\_return}_i + \beta_5 \cdot \text{mean\_rent\_royalty\_per\_rent\_royalty\_return}_i + \beta_6 \cdot \text{mean\_salary\_per\_salary\_return}_i + \beta_7 \cdot \text{total\_net\_investment\_income\_tax\_amount}_i + \beta_8 \cdot \text{mean\_tax\_payment\_per\_return}_i + \epsilon_i $$

**Model 8 (Log Income + Polynomial + Wealth):**
The mean charitable contribution per return is modeled as a function of the natural logarithm of the mean AGI per return (linear and quadratic terms) and the wealth/investment income variables, plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \ln(\text{mean\_agi\_per\_return}_i) + \beta_2 \cdot (\ln(\text{mean\_agi\_per\_return}_i))^2 + \beta_3 \cdot \text{mean\_dividend\_per\_dividend\_return}_i + \beta_4 \cdot \text{mean\_capgain\_per\_capgain\_return}_i + \beta_5 \cdot \text{mean\_rent\_royalty\_per\_rent\_royalty\_return}_i + \epsilon_i $$

**Model 9 (Log Income + Polynomial + All Other Controls):**
The mean charitable contribution per return is modeled as a function of the natural logarithm of the mean AGI per return (linear and quadratic terms) and all other control variables (wealth/investment income, salary, and tax variables), plus an error term.
Formula: $$ \text{mean\_charitable\_contributation\_per\_return}_i = \beta_0 + \beta_1 \cdot \ln(\text{mean\_agi\_per\_return}_i) + \beta_2 \cdot (\ln(\text{mean\_agi\_per\_return}_i))^2 + \beta_3 \cdot \text{mean\_dividend\_per\_dividend\_return}_i + \beta_4 \cdot \text{mean\_capgain\_per\_capgain\_return}_i + \beta_5 \cdot \text{mean\_rent\_royalty\_per\_rent\_royalty\_return}_i + \beta_6 \cdot \text{mean\_salary\_per\_salary\_return}_i + \beta_7 \cdot \text{total\_net\_investment\_income\_tax\_amount}_i + \beta_8 \cdot \text{mean\_tax\_payment\_per\_return}_i + \epsilon_i $$



### OLS Assumptions


#### IID





#### BLP Exist





### Data Splitting

```{r include=FALSE}
# Set a seed for reproducibility
set.seed(123)

# Determine the number of rows for the training set (80% of the data)
n_rows <- nrow(df)
train_size <- floor(0.8 * n_rows)

# Randomly sample indices for the training set
train_indices <- sample(1:n_rows, size = train_size, replace = FALSE)

# Create the training and testing data frames
train_df <- df[train_indices, ]
test_df <- df[-train_indices, ]
```

### Model Building

```{r include=FALSE}
# Model 1: Simple OLS (Income)
model1 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return, data = train_df)
robust_summary_model1 <- coeftest(model1, vcov = vcovHC(model1, type = "HC3"))
print("Summary of Model 1 (Simple OLS with Robust SE):")
print(summary(model1))
print("\nCoefficients (with Robust Standard Errors) - Model 1:")
print(robust_summary_model1)
cat("\n")
```

```{r include=FALSE}
# Model 2: Polynomial Income (Quadratic)
model2 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2), data = train_df)
robust_summary_model2 <- coeftest(model2, vcov = vcovHC(model2, type = "HC3"))
print("Summary of Model 2 (Polynomial Income with Robust SE):")
print(summary(model2))
print("\nCoefficients (with Robust Standard Errors) - Model 2:")
print(robust_summary_model2)
cat("\n")
```

```{r include=FALSE}
# Model 3: Log Income with Quadratic Term
model3 <- lm(mean_charitable_contributation_per_return ~ log(mean_agi_per_return) + I(log(mean_agi_per_return)^2), data = train_df)
robust_summary_model3 <- coeftest(model3, vcov = vcovHC(model3, type = "HC3"))
print("Summary of Model 3 (Log Income with Quadratic Term with Robust SE):")
print(summary(model3))
print("\nCoefficients (with Robust Standard Errors) - Model 3:")
print(robust_summary_model3)
cat("\n")
```

```{r include=FALSE}
# Model 4: Adding Wealth/Investment Income
model4 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return, data = train_df)
robust_summary_model4 <- coeftest(model4, vcov = vcovHC(model4, type = "HC3"))
print("Summary of Model 4 (Adding Wealth/Investment Income with Robust SE):")
print(summary(model4))
print("\nCoefficients (with Robust Standard Errors) - Model 4:")
print(robust_summary_model4)
cat("\n")
```

```{r include=FALSE}
# Model 5: Adding Stable Income (Salary)
model5 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return, data = train_df)
robust_summary_model5 <- coeftest(model5, vcov = vcovHC(model5, type = "HC3"))
print("Summary of Model 5 (Adding Stable Income with Robust SE):")
print(summary(model5))
print("\nCoefficients (with Robust Standard Errors) - Model 5:")
print(robust_summary_model5)
cat("\n")
```

```{r include=FALSE}
# Model 6: Adding Tax-Related Variable (Investment Income Tax)
model6 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return + total_net_investment_income_tax_amount, data = train_df)
robust_summary_model6 <- coeftest(model6, vcov = vcovHC(model6, type = "HC3"))
print("Summary of Model 6 (Adding Investment Income Tax with Robust SE):")
print(summary(model6))
print("\nCoefficients (with Robust Standard Errors) - Model 6:")
print(robust_summary_model6)
cat("\n")
```

```{r include=FALSE}
# Model 7: Adding Another Tax-Related Variable (Tax Payment)
model7 <- lm(mean_charitable_contributation_per_return ~ mean_agi_per_return + I(mean_agi_per_return^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return + total_net_investment_income_tax_amount + mean_tax_payment_per_return, data = train_df)
robust_summary_model7 <- coeftest(model7, vcov = vcovHC(model7, type = "HC3"))
print("Summary of Model 7 (Adding Tax Payment with Robust SE):")
print(summary(model7))
print("\nCoefficients (with Robust Standard Errors) - Model 7:")
print(robust_summary_model7)
cat("\n")
```

```{r include=FALSE}
# Model 8: Log Income with Quadratic Term and Wealth/Investment Income
model8 <- lm(mean_charitable_contributation_per_return ~ log(mean_agi_per_return) + I(log(mean_agi_per_return)^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return, data = train_df)
robust_summary_model8 <- coeftest(model8, vcov = vcovHC(model8, type = "HC3"))
print("Summary of Model 8 (Log Income and Wealth/Investment Income with Robust SE):")
print(summary(model8))
print("\nCoefficients (with Robust Standard Errors) - Model 8:")
print(robust_summary_model8)
cat("\n")
```

```{r include=FALSE}
# Model 9: Log Income with Quadratic Term and All Other Variables
model9 <- lm(mean_charitable_contributation_per_return ~ log(mean_agi_per_return) + I(log(mean_agi_per_return)^2) + mean_dividend_per_dividend_return + mean_capgain_per_capgain_return + mean_rent_royalty_per_rent_royalty_return + mean_salary_per_salary_return + total_net_investment_income_tax_amount + mean_tax_payment_per_return, data = train_df)
robust_summary_model9 <- coeftest(model9, vcov = vcovHC(model9, type = "HC3"))
print("Summary of Model 9 (Log Income and All Other Variables with Robust SE):")
print(summary(model9))
print("\nCoefficients (with Robust Standard Errors) - Model 9:")
print(robust_summary_model9)
cat("\n")
```

### Models Validations

```{r include=FALSE}
cat("\n--- Model Validation on the Testing Data (20%) ---\n")

# Function to evaluate model performance
evaluate_model <- function(model, test_data, true_values) {
  predictions <- predict(model, newdata = test_data)
  mse_val <- mse(true_values, predictions)
  rmse_val <- rmse(true_values, predictions)
  mae_val <- mae(true_values, predictions)
  r_squared_val <- cor(true_values, predictions)^2
  return(data.frame(MSE = mse_val, RMSE = rmse_val, MAE = mae_val, R_Squared = r_squared_val))
}

# Evaluate each model on the test data
model1_eval <- evaluate_model(model1, test_df, test_df$mean_charitable_contributation_per_return)
model2_eval <- evaluate_model(model2, test_df, test_df$mean_charitable_contributation_per_return)
model3_eval <- evaluate_model(model3, test_df, test_df$mean_charitable_contributation_per_return)
model4_eval <- evaluate_model(model4, test_df, test_df$mean_charitable_contributation_per_return)
model5_eval <- evaluate_model(model5, test_df, test_df$mean_charitable_contributation_per_return)
model6_eval <- evaluate_model(model6, test_df, test_df$mean_charitable_contributation_per_return)
model7_eval <- evaluate_model(model7, test_df, test_df$mean_charitable_contributation_per_return)
model8_eval <- evaluate_model(model8, test_df, test_df$mean_charitable_contributation_per_return)
model9_eval <- evaluate_model(model9, test_df, test_df$mean_charitable_contributation_per_return)


# Display the evaluation metrics
validation_results <- data.frame(
  Model = paste0("Model ", 1:9),
  MSE = c(model1_eval$MSE, model2_eval$MSE, model3_eval$MSE, model4_eval$MSE, model5_eval$MSE, model6_eval$MSE, model7_eval$MSE, model8_eval$MSE, model9_eval$MSE),
  RMSE = c(model1_eval$RMSE, model2_eval$RMSE, model3_eval$RMSE, model4_eval$RMSE, model5_eval$RMSE, model6_eval$RMSE, model7_eval$RMSE, model8_eval$RMSE, model9_eval$RMSE),
  MAE = c(model1_eval$MAE, model2_eval$MAE, model3_eval$MAE, model4_eval$MAE, model5_eval$MAE, model6_eval$MAE, model7_eval$MAE, model8_eval$MAE, model9_eval$MAE),
  R_Squared = c(model1_eval$R_Squared, model2_eval$R_Squared, model3_eval$R_Squared, model4_eval$R_Squared, model5_eval$R_Squared, model6_eval$R_Squared, model7_eval$R_Squared, model8_eval$R_Squared, model9_eval$R_Squared)
)

print(validation_results)
```

### Results & Discussion



The results from our regression modeling and validation strongly support our initial hypothesis of a non-linear relationship between income (specifically mean AGI) and charitable giving. The consistently significant and positive coefficient of the squared mean AGI term across several well-performing models (particularly Models 2, 4, 5, 6, and 7) indicates an accelerating rate of charitable contributions as income increases. Beyond income, wealth indicators like mean dividend and capital gains per return positively influenced giving, while total net investment income tax amount showed a negative association. Ultimately, models accounting for this non-linearity in the income-giving relationship, along with wealth and tax factors, demonstrated the most robust predictive performance on unseen data, affirming the importance of considering non-linear effects when analyzing charitable behavior.


Based on the evaluation metrics on the unseen 20% testing data, Models 6 and 7 appear to be the best performing models. They consistently show the lowest MSE, RMSE, and MAE, and the highest R-squared values on the data the models have not seen before. This indicates that these models generalize well to new data and provide the most accurate predictions for mean charitable contributions per return.

While Model 5 also performed well, Models 6 and 7 outperform it slightly in terms of overall error metrics on the test set.

The inclusion of the total_net_investment_income_tax_amount in Model 6 and the addition of mean_tax_payment_per_return in Model 7 seem to contribute to better accuracy. However, the very similar performance of Models 6 and 7 on the test set suggests that mean_tax_payment_per_return might not be adding substantial predictive power beyond what is already captured by the other variables in Model 6.
