---
title: "Intro and Methodology"
author: "Colin Frishberg"
output: pdf_document
date: "2025-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

#Introduction

Are Americans with higher incomes more generous than Americans with lower incomes? Giving USA’s Annual Report on Philanthropy for the Year 2023 reported that individual donors gave an estimated \$374.4 billion in charitable contributions, accounting for 67% of a total of $557.16 billion given [1]. With such a significant economic impact, a substantial body of research has sought to understand the intricacies of the relationship between an individual and their propensity to donate. Studies have explored the impact of various socioeconomic factors on charitable giving such as income inequality [2], underlying motivations such as altruism, warm glow or social pressure [3] and even suggested philanthropy may be having a reverse effect on wealth redistribution [4]. As income inequality over the last half-century has risen dramatically across the world, and particularly in the US over the past half century [5] a growing body of work has emerged to describe the relationship between income and donations. Despite a unanimous consensus among researchers that the amount donated increases with individuals’ incomes, existing research reveals inconsistent findings on how the proportion of income donated is related to income as highlighted in a recent literature review of 26 studies conducted on this question [6].  As Berkeley students, we are motivated by a “commitment to contributing to a better world” [7], and by providing our findings to this question we hope to add to an increasingly salient corpus of philanthropic research on income related giving behavior to be used by economists and policy makers seeking to mediate  increasing wealth inequality.

Drawing from Neumayr & Pennerstorfer’s findings on the generosity of Americans [6], we hypothesize that the relationship between income and proportion of income donated is non-linear. Specifically, we predict a U-shaped relationship where both the lowest and highest income Americans donate a larger proportion of their income than median income Americans. To model this predicted parabolic relationship, we will use an OLS model keeping our parameters linear while applying a logarithmic transformation on income and its quadratic term to model our estimated relationship. 

#Data Source and Methodology

The data source we have chosen to use is the IRS’ Statistics of Income (SOI) program’s ZIP Code data for the tax year 2022. This data set aggregates (at the ZIP code level)  individual income tax returns ( with variables assigned to the various items in the 1040 form) filed with the Internal Revenue Service (IRS) during the 12-month period of January 1, 2023 to December 31, 2023. There are 27,690 observations of 165 variables which the IRS samples from the master file system containing the over 200 million tax returns collected each year, compiling statistics based on stratified probability samples of tax or information returns [8]. To normalize the data and avoid the common pitfall of using ZIP codes for analysis [9] we’ve chosen to merge the ZIP codes with US Census Bureau county geographies using the Housing and Urban Development’s  HUD USPS -ZIP Crosswalk file for the 4th quarter of 2022 [10]. We've also calculated proportional statistics such as agi per return etc. as a method of standardization. Since we are interested in individuals, and not businesses, we will use the provided residential ratio (from HUD file) to allocate ZIP code level data to residential addresses taking into account both the spatial distribution of population, and the spatial distribution of residences. These normalization steps and some additional data processing found in the data_wrangling.R file are used to produce the data set we leverage in our analysis. 

We recognize the following important limitations of our analysis due to limitations in the underlying data.

1. The IRS Data does not represent the full U.S. population because many individuals are not required to file an individual income tax return. Therefore our results will not be appropriate for generalizations of the US population and we may be excluding a significant number of very low income individuals who donate to charity. 
2. Due to anonymity measures taken by the IRS, some US zip codes will not be represented in our study so our results equally will not be appropriate for ZIP code generalization. We omit these ZIP codes as part of our analysis which is a reduction of 102 rows. 
3. Although we take many steps to normalize the data, we do acknowledge that by using aggregated data we do run a risk of ecological fallacy.
4. Poorer households are more likely to take the standard deduction rather than itemise charitable donations [6] which means that charitable giving is likely to be under-reported for lower income groups.

# Data Wrangling

## Load Data Sets
```{r , echo =FALSE }
#Load IRS 2022 ZipCode level Statistics of Income data set
irs <- read_csv("irs_no_agi_raw.csv")
#Load HUD Crosswalk file which maps Zips -> Counties including allocation weights
hud <- readxl::read_xlsx("zip_county_122022.xlsx")

```

##Filter for US States
```{r, echo =FALSE}
#filter out US territories from HUD data set
hud <- hud |>
  filter(!USPS_ZIP_PREF_STATE %in% c("AS", "GU", "MP", "PR", "VI"))

```



##Prep for Join
```{r pressure, echo=FALSE}
#rename columns to facilitate join with irs
hud <- hud |>
  rename(ZIPCODE = ZIP, county_fips = COUNTY) |>
  rename_all(tolower) |>
  mutate(state_fips_hud = substr(county_fips, 1, 2))

#check how many Zips span multiple counties (require allocation after join)
hud |>
  group_by(zipcode) |>
  summarise(county_count = n_distinct(county_fips)) |>
  filter(county_count > 1) |>
  arrange(desc(county_count))

#looks like 11K one to many zip -> county allocations we need to perform

#clean up irs keeping target variables
irs <- irs |>
  mutate(
    zipcode = ZIPCODE,
    state_fips_irs = STATEFIPS,
    total_charitable_contb = A19700,
    agi = A00100,
    num_of_returns = N1,
    net_investment_income_tax = A85300,
    net_cap_gains_less_loss = A01000,
    num_returns_net_cap_gains_less_loss = N01000,
    num_returns_rent_and_royalty = N25870,
    total_rent_and_royalty = A25870,
    ord_div_amount = A00600,
    num_returns_ord_dividens = N00600,
    state = STATE,
    total_tax_payments_amount = A10600,
    salaries_and_wages_amnt = A00200,
    num_salaries_and_wages = N00200,
    zip_region = substr(ZIPCODE, 1, 3),
    .keep = "none"
  )  |>
  filter(!zipcode %in% c('00000', '99999')) 

```

## Join Data Sets and perform checks
```{r}
#join
county_level_irs_data <- left_join(irs, hud, by = "zipcode")

# check for NAs
county_level_irs_data |> filter(is.na(county_fips)) #none that is good

#check allocation targets again, slightly less at 10.6K
county_level_irs_data |>
  group_by(zipcode) |>
  summarise(county_count = n_distinct(county_fips)) |>
  filter(county_count > 1) |>
  arrange(desc(county_count))

#sense check comparing state_fips between each data source

county_level_irs_data <- county_level_irs_data |>
  mutate(
    state_fips_match_strict = case_when(
      # Both NA -> considered a match here
      is.na(state_fips_irs) & is.na(state_fips_hud) ~ TRUE,
      # One NA, one not -> Mismatch
      is.na(state_fips_irs) | is.na(state_fips_hud) ~ FALSE,
      # Both have values and are equal -> Match
      state_fips_irs == state_fips_hud            ~ TRUE,
      # Both have values and are unequal -> Mismatch
      TRUE                                        ~ FALSE
    )
  )

# Count the explicit matches and mismatches
print(county_level_irs_data |> count(state_fips_match_strict))

# Count NAs
print(county_level_irs_data |> count(is.na(state_fips_irs), is.na(state_fips_hud)))

# View only the rows definitively flagged as mismatches
print(
  county_level_irs_data |>
    filter(!state_fips_match_strict) |>
    select(zipcode, county_fips, state_fips_irs, state_fips_hud)
)

#Looks like an error in the HUD crosswalk file where ZIP 20736 is being
#attributed to Yuma County AZ but should be in MD. The other mismatch is expected
#based on methodology of HUD Crosswalk 

# filtering out that row before allocation and aggregation
combined_irs_hud_filtered <- county_level_irs_data |>
  filter(!(zipcode == "20736" & county_fips == "04025"))

#Select rows we are interested in, and reorganize chr vars to front of DF

combined_irs_hud_filtered <- combined_irs_hud_filtered |>
  select(!state_fips_match_strict)

combined_irs_hud_filtered <- combined_irs_hud_filtered |>
  relocate(where(is.character), .before = 1)

# Final Output 
glimpse(combined_irs_hud_filtered) # = interim_irs_hud_combined.csv

```

## Apply Allocations
```{r}
## Step 2: Now we make use of the allocation ratio columns (specifically RES_RATIO)

allocated_data <- combined_irs_hud_filtered |>
  mutate(# Multiply each IRS variable by the Residential allocation ratio
    across(
      c(
        agi,
        num_of_returns,
        total_charitable_contb,
        net_investment_income_tax,
        net_cap_gains_less_loss,
        num_returns_net_cap_gains_less_loss,
        num_returns_rent_and_royalty,
        total_rent_and_royalty,
        ord_div_amount,
        num_returns_ord_dividens,
        total_tax_payments_amount,
        salaries_and_wages_amnt,
        num_salaries_and_wages
      ),
      ~ . * res_ratio,
      .names = "{.col}_allocated"
    ))

```

## Aggregate at County Level

```{r}
#Now we aggregate at county level


county_level_irs_data <- allocated_data |>
  # Group by the unique county identifier
  group_by(county_fips) |>
  # Summarise by summing the allocated values for each county
  summarise(
    # Count and Main Predictor
    total_returns = sum(num_of_returns_allocated, na.rm = TRUE),
    total_agi_amount = sum(agi_allocated, na.rm = TRUE),
    
    # Salaries and Wages
    total_salaries_wages_amount = sum(salaries_and_wages_amnt_allocated, na.rm = TRUE),
    total_salaries_wages_returns = sum(num_salaries_and_wages_allocated, na.rm = TRUE),
    
    # Ordinary Dividends
    total_ordinary_dividends_amount = sum(ord_div_amount_allocated, na.rm = TRUE),
    total_ordinary_dividends_returns = sum(num_returns_ord_dividens_allocated, na.rm = TRUE),
    
    # Net Capital Gains less Loss
    total_net_capital_gains_amount = sum(net_cap_gains_less_loss_allocated, na.rm = TRUE),
    total_net_capital_gains_returns = sum(num_returns_net_cap_gains_less_loss_allocated, na.rm = TRUE),
    
    # (Outcome) Charitable Contributions
    total_charitable_contributions_amount = sum(total_charitable_contb_allocated, na.rm = TRUE),
    
    # Net Investment Income Tax
    total_net_investment_income_tax_amount = sum(net_investment_income_tax_allocated, na.rm = TRUE),
    
    # Rent and Royalty Income
    total_rent_royalty_amount = sum(total_rent_and_royalty_allocated, na.rm = TRUE),
    total_rent_royalty_returns = sum(num_returns_rent_and_royalty_allocated, na.rm = TRUE),
    
    # Total Tax Payments
    total_tax_payments_amount = sum(total_tax_payments_amount_allocated, na.rm = TRUE),
    
    # Keep the state FIPS (should be unique per county_fips)
    state_fips = first(state_fips_hud),
    # Use the state FIPS from the HUD data context
    
    .groups = 'drop' # Ungroup after summarising
  ) 

```

## Proportion Normalization
```{r}
#Proportion Normalization
county_level_irs_data <- county_level_irs_data |>
  mutate(
    #Total return normalizations
    mean_agi_per_return = ifelse(total_returns > 0, total_agi_amount / total_returns, 0),
    mean_charitable_contributation_per_return = ifelse(
      total_returns > 0,
      total_charitable_contributions_amount / total_returns,
      0
    ),
    mean_tax_payment_per_return = ifelse(total_returns > 0, total_tax_payments_amount / total_returns, 0),
    
    #Specific return normalizations
    mean_salary_per_salary_return = ifelse(
      total_salaries_wages_returns > 0,
      total_salaries_wages_amount / total_salaries_wages_returns,
      0
    ),
    
    
    mean_dividend_per_dividend_return = ifelse(
      total_ordinary_dividends_returns > 0,
      total_ordinary_dividends_amount / total_ordinary_dividends_returns,
      0
    ),
    
    
    mean_capgain_per_capgain_return = ifelse(
      total_net_capital_gains_returns > 0,
      total_net_capital_gains_amount / total_net_capital_gains_returns,
      0
    ),
    
    
    mean_rent_royalty_per_rent_royalty_return = ifelse(
      total_rent_royalty_returns > 0,
      total_rent_royalty_amount / total_rent_royalty_returns,
      0
    )
  )

#Final output 
glimpse(county_level_irs_data) # = "interim_irs_county.csv"
```

